maze-dataset: Maze Generation with Algorithmic Variety and Representational Flexibility

### 2025-07-28 docs/paper/paper.tex

\documentclass[10pt,a4paper,onecolumn]{article}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage[rgb,svgnames]{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{tikz}
\usepackage[pdfa]{hyperref}
\usepackage{hyperxmp}
\hypersetup{%
    unicode=true,
    pdfapart=3,
    pdfaconformance=B,
    pdftitle={maze-dataset: Maze Generation with Algorithmic Variety and
Representational Flexibility},
    pdfauthor={Michael Igorevich Ivanitskiy, Aaron Sandoval, Alex F.
Spies, Tilman Räuker, Brandon Knutson, Cecilia Diniz Behn, Samy Wu
Fung},
    pdfpublication={Journal of Open Source Software},
    pdfpublisher={Open Journals},
    pdfissn={2475-9066},
    pdfpubtype={journal},
    pdfvolumenum={},
    pdfissuenum={},
    pdfdoi={N/A},
    pdfcopyright={Copyright (c) 1970, Michael Igorevich Ivanitskiy,
Aaron Sandoval, Alex F. Spies, Tilman Räuker, Brandon Knutson, Cecilia
Diniz Behn, Samy Wu Fung},
    pdflicenseurl={http://creativecommons.org/licenses/by/4.0/},
    colorlinks=true,
    linkcolor=[rgb]{0.0, 0.5, 1.0},
    citecolor=Blue,
    urlcolor=[rgb]{0.0, 0.5, 1.0},
    breaklinks=true
}
% https://tex.stackexchange.com/a/535849
% Create an OutputIntent in order to correctly specify colours
\immediate\pdfobj stream attr{/N 3} file{sRGB.icc}
\pdfcatalog{%
  /OutputIntents [
    <<
      /Type /OutputIntent
      /S /GTS_PDFA1
      /DestOutputProfile \the\pdflastobj\space 0 R
      /OutputConditionIdentifier (sRGB)
      /Info (sRGB)
    >>
  ]
}
\pdfvariable omitcidset=1
\usepackage{caption}
\usepackage{orcidlink}
\usepackage{tcolorbox}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{seqsplit}
\usepackage{xstring}

@@

\fancyfoot[L]{\parbox[t]{0.98\headwidth}{\footnotesize{\sffamily Ivanitskiy
et al. (1970). maze-dataset: Maze Generation with Algorithmic Variety
and Representational Flexibility. \emph{Journal of Open Source
Software}, \emph{¿VOL?}(¿ISSUE?), ¿PAGE? \url{https://doi.org/N/A}}.}}

@@

\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\urlstyle{same}  % don't use monospace font for urls
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}

@@

  {\bfseries DOI:} \href{https://doi.org/N/A}{\color{linky}{N/A}}

@@

    {\bfseries Submitted:} 01 January 1970\\
    {\bfseries Published:} 01 January 1970

@@

\hypertarget{summary}{%
\section{Summary}\label{summary}}

@@

\hypertarget{statement-of-need}{%
\section{Statement of Need}\label{statement-of-need}}

@@

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survey}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we found that
existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) lack support for transforming between multiple
representations and provide limited control over the maze generation
process.

\hypertarget{related-works}{%
\subsection{Related Works}\label{related-works}}

A multitude of public and open-source software packages exist for
generating mazes (\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). However, nearly all of these packages produce
mazes represented as rasterized images or other visual formats rather
than the underlying graph structure, and this makes it difficult to work
with these datasets.

@@

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}
    (\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
    Bansal, et al., 2021}) and can be configured to be similar to the
    outputs of Németh (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature‑rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\protect\hyperlink{ref-mazegenerator-net}{Alance AB,
    2019}; \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \&
  Wu, 2023}) but with many more options, detailed
  \secref{sec:tokenized-output-formats}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in \secref{sec:implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

\hypertarget{features}{%
\section{Features}\label{features}}

We direct readers to our
\href{https://understanding-search.github.io/maze-dataset/examples/maze_examples.html}{examples},
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html}{docs},
and
\href{https://understanding-search.github.io/maze-dataset/notebooks/}{notebooks}
for more information. Our package can be installed from
\href{https://pypi.org/project/maze-dataset/}{PyPi} via
\texttt{pip\ install\ maze-dataset}, or directly from the
\href{https://github.com/understanding-search/maze-dataset}{git
repository} (\protect\hyperlink{ref-maze-dataset-github}{Michael I.
Ivanitskiy et al., 2023a}).

Datasets of mazes are created from a
\docslinkcodemain{MazeDatasetConfig} configuration object, which allows
specifying the number of mazes, their size, the generation algorithm,
and various parameters for the generation algorithm. Datasets can also
be filtered after generation to satisfy certain properties. Custom
filters can be specified, and some filters are included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}

\hypertarget{visual-output-formats}{%
\subsection{Visual Output Formats}\label{visual-output-formats}}

@@

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\protect\hyperlink{ref-deepthinking}{Schwarzschild, Borgnia, Gupta,
Huang, et al., 2021}). To facilitate the use of our package in this
context, we replicate the format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) and provide the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}}
class which returns rasterized pairs of (input, target) mazes as shown
in \autoref{fig:e2h-raster}.

@@

\hypertarget{sec:tokenized-output-formats}{%
\subsection{Tokenized Output
Formats}\label{sec:tokenized-output-formats}}

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\protect\hyperlink{ref-pfau2024dotbydot}{Pfau et al.,
2024}; \protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al.,
2024}). To facilitate systematic investigation of the effects of
different representations of data on text model performance, we provide
a variety of text output formats, with an example given in
\autoref{fig:token-regions}.

@@

\hypertarget{benchmarks}{%
\subsection{Benchmarks}\label{benchmarks}}

@@

\hypertarget{sec:implementation}{%
\section{Implementation}\label{sec:implementation}}

@@

\hypertarget{usage-in-research}{%
\section{Usage in Research}\label{usage-in-research}}

This package was originally built for the needs of the
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}) project, which aims to investigate spatial planning and
world models in autoregressive transformer models trained on mazes
(\protect\hyperlink{ref-ivanitskiy2023structuredworldreps}{Michael
Igorevich Ivanitskiy, Spies, et al., 2023};
\protect\hyperlink{ref-maze-dataset-arxiv-2023}{Michael Igorevich
Ivanitskiy, Shah, et al., 2023};
\protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al., 2024}).
It was extended for work on understanding the mechanisms by which
recurrent convolutional and implicit networks
(\protect\hyperlink{ref-fung2022jfb}{Fung et al., 2022}) solve mazes
given a rasterized view
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), which required matching the pixel-padded and endpoint
constrained output format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}). Ongoing work using \texttt{maze-dataset} aims to
investigate the effects of varying the tokenization format on the
performance of pretrained LLMs on spatial reasoning.

@@

\begin{itemize}
\item
  By (\protect\hyperlink{ref-nolte2024multistep}{Nolte et al., 2024}) to
  compare the effectiveness of transformers trained with the
  MLM-\(\mathcal{U}\)
  (\protect\hyperlink{ref-MLMU-kitouni2024factorization}{Kitouni et al.,
  2024}) multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  By (\protect\hyperlink{ref-wang2024imperative}{Wang et al., 2024}) and
  (\protect\hyperlink{ref-chen2024iaimperative}{Chen et al., 2024}) to
  study the effectiveness of imperative learning.
\item
  By (\protect\hyperlink{ref-zhang2025tscend}{Zhang et al., 2025}) to
  introduce a novel framework for reasoning diffusion models.
\item
  By (\protect\hyperlink{ref-dao2025alphamaze}{Dao \& Vu, 2025}) to
  improve spatial reasoning in LLMs with GRPO.
\end{itemize}

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0.5}
\leavevmode\vadjust pre{\hypertarget{ref-mazegenerator-net}{}}%
Alance AB. (2019). \emph{Maze generator}.
\url{http://www.mazegenerator.net}.

\leavevmode\vadjust pre{\hypertarget{ref-ayaz2008maze}{}}%
Ayaz, H., Allen, S. L., Platek, S. M., \& Onaral, B. (2008). Maze suite
1.0: A complete set of tools to prepare, present, and analyze
navigational and spatial cognitive neuroscience experiments.
\emph{Behavior Research Methods}, \emph{40}, 353--359.
\url{https://doi.org/10.3758/brm.40.1.353}

\leavevmode\vadjust pre{\hypertarget{ref-chen2024iaimperative}{}}%
Chen, X., Yang, F., \& Wang, C. (2024). {iA*}: Imperative learning-based
{A*} search for pathfinding. \emph{arXiv Preprint arXiv:2403.15870}.
\url{https://doi.org/10.48550/arXiv.2403.15870}

\leavevmode\vadjust pre{\hypertarget{ref-cobbe2019procgen}{}}%
Cobbe, K., Hesse, C., Hilton, J., \& Schulman, J. (2019). Leveraging
procedural generation to benchmark reinforcement learning. \emph{arXiv
Preprint arXiv:1912.01588}.
\url{https://doi.org/10.48550/arXiv.1912.01588}

\leavevmode\vadjust pre{\hypertarget{ref-dao2025alphamaze}{}}%
Dao, A., \& Vu, D. B. (2025). AlphaMaze: Enhancing large language
models' spatial intelligence via GRPO. \emph{arXiv Preprint
arXiv:2502.14669}. \url{https://doi.org/10.48550/arXiv.2502.14669}

\leavevmode\vadjust pre{\hypertarget{ref-gh_Ehsan_2022}{}}%
Ehsan, E. (2022). \emph{Maze}. \url{https://github.com/emadehsan/maze}

\leavevmode\vadjust pre{\hypertarget{ref-fung2022jfb}{}}%
Fung, S. W., Heaton, H., Li, Q., McKenzie, D., Osher, S., \& Yin, W.
(2022). Jfb: Jacobian-free backpropagation for implicit networks.
\emph{Proceedings of the AAAI Conference on Artificial Intelligence},
\emph{36}, 6648--6656. \url{https://doi.org/10.1609/aaai.v36i6.20619}

\leavevmode\vadjust pre{\hypertarget{ref-mathematica-maze}{}}%
Guo, C., Barthelet, L., \& Morris, R. (2011). \emph{Maze generator and
solver}. Wolfram Demonstrations Project,
\url{https://demonstrations.wolfram.com/MazeGeneratorAndSolver/}.

\leavevmode\vadjust pre{\hypertarget{ref-harriesMazeExplorerCustomisable3D2019}{}}%
Harries, L., Lee, S., Rzepecki, J., Hofmann, K., \& Devlin, S. (n.d.).
{MazeExplorer}: {A Customisable 3D Benchmark} for {Assessing
Generalisation} in {Reinforcement Learning}. \emph{2019 {IEEE Conf}.
{Games CoG}}, 1--4. \url{https://doi.org/10.1109/cig.2019.8848048}

\leavevmode\vadjust pre{\hypertarget{ref-zanj}{}}%
Ivanitskiy, M. (n.d.). \emph{ZANJ}.
\url{https://doi.org/10.5281/zenodo.15540393}

\leavevmode\vadjust pre{\hypertarget{ref-maze-dataset-github}{}}%
Ivanitskiy, Michael I., Shah, R., Spies, A. F., Räuker, T., Valentine,
D., Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023a).
\emph{Maze dataset}. \url{https://doi.org/10.48550/arXiv.2309.10498}

\leavevmode\vadjust pre{\hypertarget{ref-maze-transformer-github}{}}%
Ivanitskiy, Michael I., Shah, R., Spies, A. F., Räuker, T., Valentine,
D., Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023b).
\emph{Maze transformer interpretability}.
\url{https://doi.org/10.48550/arXiv.2312.02566}

\leavevmode\vadjust pre{\hypertarget{ref-maze-dataset-arxiv-2023}{}}%
Ivanitskiy, Michael Igorevich, Shah, R., Spies, A. F., Räuker, T.,
Valentine, D., Rager, C., Quirke, L., Mathwin, C., Corlouer, G., Behn,
C. D., \& others. (2023). A configurable library for generating and
manipulating maze datasets. \emph{arXiv Preprint arXiv:2309.10498}.
\url{https://doi.org/10.48550/arXiv.2309.10498}

\leavevmode\vadjust pre{\hypertarget{ref-ivanitskiy2023structuredworldreps}{}}%
Ivanitskiy, Michael Igorevich, Spies, A. F., Räuker, T., Corlouer, G.,
Mathwin, C., Quirke, L., Rager, C., Shah, R., Valentine, D., Behn, C.
D., \& others. (2023). Structured world representations in maze-solving
transformers. \emph{arXiv Preprint arXiv:2312.02566}.
\url{https://doi.org/10.48550/arXiv.2312.02566}

>>>>

\documentclass[10pt,a4paper,onecolumn]{article}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage[rgb,svgnames]{xcolor}
\usepackage{authblk,etoolbox}
\usepackage{titlesec}
\usepackage{calc}
\usepackage{tikz}
\usepackage[pdfa]{hyperref}
\usepackage{hyperxmp}
\hypersetup{%
    unicode=true,
    pdfapart=3,
    pdfaconformance=B,
    pdftitle={maze-dataset: Maze Generation with Algorithmic Variety and
Representational Flexibility},
    pdfauthor={Michael Igorevich Ivanitskiy, Aaron Sandoval, Alex F.
Spies, Tilman Räuker, Brandon Knutson, Cecilia Diniz Behn, Samy Wu
Fung},
    pdfpublication={Journal of Open Source Software},
    pdfpublisher={Open Journals},
    pdfissn={2475-9066},
    pdfpubtype={journal},
    pdfvolumenum={},
    pdfissuenum={},
    pdfdoi={10.xxxxxx/draft},
    pdfcopyright={Copyright (c) 1970, Michael Igorevich Ivanitskiy,
Aaron Sandoval, Alex F. Spies, Tilman Räuker, Brandon Knutson, Cecilia
Diniz Behn, Samy Wu Fung},
    pdflicenseurl={http://creativecommons.org/licenses/by/4.0/},
    colorlinks=true,
    linkcolor=[rgb]{0.0, 0.5, 1.0},
    citecolor=Blue,
    urlcolor=[rgb]{0.0, 0.5, 1.0},
    breaklinks=true
}
% https://tex.stackexchange.com/a/535849
% Create an OutputIntent in order to correctly specify colours
\immediate\pdfobj stream attr{/N 3} file{sRGB.icc}
\pdfcatalog{%
  /OutputIntents [
    <<
      /Type /OutputIntent
      /S /GTS_PDFA1
      /DestOutputProfile \the\pdflastobj\space 0 R
      /OutputConditionIdentifier (sRGB)
      /Info (sRGB)
    >>
  ]
}
\pdfvariable omitcidset=1
\usepackage{caption}
\usepackage{orcidlink}
\usepackage{tcolorbox}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{seqsplit}
\usepackage{xstring}

@@

\fancyfoot[L]{\parbox[t]{0.98\headwidth}{\footnotesize{\sffamily Ivanitskiy
et al. (2025). maze-dataset: Maze Generation with Algorithmic Variety
and Representational Flexibility. \emph{Journal of Open Source
Software}, \emph{¿VOL?}(¿ISSUE?), ¿PAGE?
\url{https://doi.org/10.xxxxxx/draft}}.}}

@@

\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\urlstyle{same}  % don't use monospace font for urls
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{american}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\usepackage{lineno}
\linenumbers
\usepackage{draftwatermark}

@@

  {\bfseries DOI:} \href{https://doi.org/10.xxxxxx/draft}{\color{linky}{10.xxxxxx/draft}}

@@

    {\bfseries Submitted:} 01 January 1970\\
    {\bfseries Published:} unpublished

@@

\section{Summary}\label{summary}

@@

\section{Statement of Need}\label{statement-of-need}

@@

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\citeproc{ref-interpretability-survey}{Räuker et al., 2023}). Although
mazes are well suited for these investigations, we found that existing
maze generation packages (\citeproc{ref-cobbe2019procgen}{Cobbe et al.,
2019}; \citeproc{ref-gh_Ehsan_2022}{Ehsan, 2022};
\citeproc{ref-harriesMazeExplorerCustomisable3D2019}{Harries et al.,
n.d.}; \citeproc{ref-gh_Nemeth_2019}{Németh, 2019};
\citeproc{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta, Bansal, et
al., 2021}) lack support for transforming between multiple
representations and provide limited control over the maze generation
process.

\subsection{Related Works}\label{related-works}

A multitude of public and open-source software packages exist for
generating mazes (\citeproc{ref-gh_Ehsan_2022}{Ehsan, 2022};
\citeproc{ref-gh_Nemeth_2019}{Németh, 2019};
\citeproc{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta, Bansal, et
al., 2021}). However, nearly all of these packages produce mazes
represented as rasterized images or other visual formats rather than the
underlying graph structure, and this makes it difficult to work with
these datasets.

@@

  \begin{itemize}
  \tightlist
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}},
    utilizing
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_pixels}{\texttt{as\_pixels()}},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}
    (\citeproc{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta, Bansal,
    et al., 2021}) and can be configured to be similar to the outputs of
    Németh (\citeproc{ref-gh_Nemeth_2019}{2019})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.as_ascii}{\texttt{as\_ascii()}}
    provides a format similar to
    (\citeproc{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \citeproc{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \href{https://understanding-search.github.io/maze-dataset/maze_dataset/plotting.html\#MazePlot}{\texttt{MazePlot}}
    provides a feature‑rich plotting utility with support for multiple
    paths, heatmaps over positions, and more. This is similar to the
    outputs of (\citeproc{ref-mazegenerator-net}{Alance AB, 2019};
    \citeproc{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \citeproc{ref-mathematica-maze}{Guo et al., 2011};
    \citeproc{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by
  \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.as_tokens}{\texttt{SolvedMaze(...).as\_tokens()}}
  is similar to that of (\citeproc{ref-eval-LLM-graphs}{Liu \& Wu,
  2023}) but with many more options, detailed in
  \hyperref[sec:tokenized-output-formats]{section: \textit{\nameref{sec:tokenized-output-formats}}}.
\item
  Preserving metadata about the generation algorithm with the dataset
  itself is essential for studying the effects of distributional shifts.
  Our package efficiently stores the dataset along with its metadata in
  a single human-readable file (\citeproc{ref-zanj}{M. Ivanitskiy,
  n.d.}). As far as we are aware, no existing packages do this reliably.
\item
  Storing mazes as images or adjacency matrices is not only difficult to
  work with, but also inefficient. We use a highly efficient method
  detailed in
  \hyperref[sec:implementation]{section: \textit{\nameref{sec:implementation}}}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\citeproc{ref-ayaz2008maze}{Ayaz et al., 2008})
  appear to simply no longer be accessible.
\end{itemize}

@@

\section{Features}\label{features}

We direct readers to our
\href{https://understanding-search.github.io/maze-dataset/examples/maze_examples.html}{examples},
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html}{docs},
and
\href{https://understanding-search.github.io/maze-dataset/notebooks/}{notebooks}
for more information. Our package can be installed from
\href{https://pypi.org/project/maze-dataset/}{PyPi} via
\texttt{pip\ install\ maze-dataset}, or directly from the
\href{https://github.com/understanding-search/maze-dataset}{git
repository} (\citeproc{ref-maze-dataset-github}{Michael I. Ivanitskiy et
al., 2023a}).

Datasets of mazes are created from a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDatasetConfig}{\texttt{MazeDatasetConfig}}
configuration object, which allows specifying the number of mazes, their
size, the generation algorithm, and various parameters for the
generation algorithm. Datasets can also be filtered after generation to
satisfy certain properties. Custom filters can be specified, and some
filters are included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}

\subsection{Visual Output Formats}\label{visual-output-formats}

@@

In previous work, maze tasks have been used with Recurrent Convolutional
Neural Network (RCNN) derived architectures
(\citeproc{ref-deepthinking}{Schwarzschild, Borgnia, Gupta, Huang, et
al., 2021}). To facilitate the use of our package in this context, we
replicate the format of (\citeproc{ref-easy_to_hard}{Schwarzschild,
Borgnia, Gupta, Bansal, et al., 2021}) and provide the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/rasterized.html\#RasterizedMazeDataset}{\texttt{RasterizedMazeDataset}}
class which returns rasterized pairs of (input, target) mazes as shown
in \autoref{fig:e2h-raster}.

@@

\subsection{Tokenized Output
Formats}\label{sec:tokenized-output-formats}

Autoregressive transformer models can be quite sensitive to the exact
format of input data, and may even use delimiter tokens to perform
reasoning steps (\citeproc{ref-pfau2024dotbydot}{Pfau et al., 2024};
\citeproc{ref-spies2024causalworldmodels}{Spies et al., 2024}). To
facilitate systematic investigation of the effects of different
representations of data on text model performance, we provide a variety
of text output formats, with an example given in
\autoref{fig:token-regions}.

@@

\subsection{Benchmarks}\label{benchmarks}

@@

\section{Implementation}\label{sec:implementation}

@@

\section{Usage in Research}\label{usage-in-research}

This package was originally built for the needs of the
(\citeproc{ref-maze-transformer-github}{Michael I. Ivanitskiy et al.,
2023b}) project, which aims to investigate spatial planning and world
models in autoregressive transformer models trained on mazes
(\citeproc{ref-ivanitskiy2023structuredworldreps}{Michael Igorevich
Ivanitskiy, Spies, et al., 2023};
\citeproc{ref-maze-dataset-arxiv-2023}{Michael Igorevich Ivanitskiy,
Shah, et al., 2023}; \citeproc{ref-spies2024causalworldmodels}{Spies et
al., 2024}). It was extended for work on understanding the mechanisms by
which recurrent convolutional and implicit networks
(\citeproc{ref-fung2022jfb}{Fung et al., 2022}) solve mazes given a
rasterized view (\citeproc{ref-knutson2024logicalextrapolation}{Knutson
et al., 2024}), which required matching the pixel-padded and endpoint
constrained output format of (\citeproc{ref-easy_to_hard}{Schwarzschild,
Borgnia, Gupta, Bansal, et al., 2021}). Ongoing work using
\texttt{maze-dataset} aims to investigate the effects of varying the
tokenization format on the performance of pretrained LLMs on spatial
reasoning.

@@

\begin{itemize}
\item
  By (\citeproc{ref-nolte2024multistep}{Nolte et al., 2024}) to compare
  the effectiveness of transformers trained with the MLM-\(\mathcal{U}\)
  (\citeproc{ref-MLMU-kitouni2024factorization}{Kitouni et al., 2024})
  multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  By (\citeproc{ref-wang2024imperative}{Wang et al., 2024}) and
  (\citeproc{ref-chen2024iaimperative}{Chen et al., 2024}) to study the
  effectiveness of imperative learning.
\item
  By (\citeproc{ref-zhang2025tscend}{Zhang et al., 2025}) to introduce a
  novel framework for reasoning diffusion models.
\item
  By (\citeproc{ref-dao2025alphamaze}{Dao \& Vu, 2025}) to improve
  spatial reasoning in LLMs with GRPO.
\end{itemize}

\section{Acknowledgements}\label{acknowledgements}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0.5}
\bibitem[\citeproctext]{ref-mazegenerator-net}
Alance AB. (2019). \emph{Maze generator}.
\url{http://www.mazegenerator.net}.

\bibitem[\citeproctext]{ref-ayaz2008maze}
Ayaz, H., Allen, S. L., Platek, S. M., \& Onaral, B. (2008). Maze suite
1.0: A complete set of tools to prepare, present, and analyze
navigational and spatial cognitive neuroscience experiments.
\emph{Behavior Research Methods}, \emph{40}, 353--359.
\url{https://doi.org/10.3758/brm.40.1.353}

\bibitem[\citeproctext]{ref-chen2024iaimperative}
Chen, X., Yang, F., \& Wang, C. (2024). {iA*}: Imperative learning-based
{A*} search for pathfinding. \emph{arXiv Preprint arXiv:2403.15870}.
\url{https://doi.org/10.48550/arXiv.2403.15870}

\bibitem[\citeproctext]{ref-cobbe2019procgen}
Cobbe, K., Hesse, C., Hilton, J., \& Schulman, J. (2019). Leveraging
procedural generation to benchmark reinforcement learning. \emph{arXiv
Preprint arXiv:1912.01588}.
\url{https://doi.org/10.48550/arXiv.1912.01588}

\bibitem[\citeproctext]{ref-dao2025alphamaze}
Dao, A., \& Vu, D. B. (2025). AlphaMaze: Enhancing large language
models' spatial intelligence via GRPO. \emph{arXiv Preprint
arXiv:2502.14669}. \url{https://doi.org/10.48550/arXiv.2502.14669}

\bibitem[\citeproctext]{ref-gh_Ehsan_2022}
Ehsan, E. (2022). \emph{Maze}. \url{https://github.com/emadehsan/maze}

\bibitem[\citeproctext]{ref-fung2022jfb}
Fung, S. W., Heaton, H., Li, Q., McKenzie, D., Osher, S., \& Yin, W.
(2022). Jfb: Jacobian-free backpropagation for implicit networks.
\emph{Proceedings of the AAAI Conference on Artificial Intelligence},
\emph{36}, 6648--6656. \url{https://doi.org/10.1609/aaai.v36i6.20619}

\bibitem[\citeproctext]{ref-mathematica-maze}
Guo, C., Barthelet, L., \& Morris, R. (2011). \emph{Maze generator and
solver}. Wolfram Demonstrations Project,
\url{https://demonstrations.wolfram.com/MazeGeneratorAndSolver/}.

\bibitem[\citeproctext]{ref-harriesMazeExplorerCustomisable3D2019}
Harries, L., Lee, S., Rzepecki, J., Hofmann, K., \& Devlin, S. (n.d.).
{MazeExplorer}: {A Customisable 3D Benchmark} for {Assessing
Generalisation} in {Reinforcement Learning}. \emph{2019 {IEEE Conf}.
{Games CoG}}, 1--4. \url{https://doi.org/10.1109/cig.2019.8848048}

\bibitem[\citeproctext]{ref-zanj}
Ivanitskiy, M. (n.d.). \emph{ZANJ}.
\url{https://doi.org/10.5281/zenodo.15540393}

\bibitem[\citeproctext]{ref-maze-dataset-github}
Ivanitskiy, Michael I., Shah, R., Spies, A. F., Räuker, T., Valentine,
D., Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023a).
\emph{Maze dataset}. \url{https://doi.org/10.48550/arXiv.2309.10498}

\bibitem[\citeproctext]{ref-maze-transformer-github}
Ivanitskiy, Michael I., Shah, R., Spies, A. F., Räuker, T., Valentine,
D., Rager, C., Quirke, L., Corlouer, G., \& Mathwin, C. (2023b).
\emph{Maze transformer interpretability}.
\url{https://doi.org/10.48550/arXiv.2312.02566}

\bibitem[\citeproctext]{ref-maze-dataset-arxiv-2023}
Ivanitskiy, Michael Igorevich, Shah, R., Spies, A. F., Räuker, T.,
Valentine, D., Rager, C., Quirke, L., Mathwin, C., Corlouer, G., Behn,
C. D., \& others. (2023). A configurable library for generating and
manipulating maze datasets. \emph{arXiv Preprint arXiv:2309.10498}.
\url{https://doi.org/10.48550/arXiv.2309.10498}

\bibitem[\citeproctext]{ref-ivanitskiy2023structuredworldreps}
Ivanitskiy, Michael Igorevich, Spies, A. F., Räuker, T., Corlouer, G.,
Mathwin, C., Quirke, L., Rager, C., Shah, R., Valentine, D., Behn, C.
D., \& others. (2023). Structured world representations in maze-solving
transformers. \emph{arXiv Preprint arXiv:2312.02566}.
\url{https://doi.org/10.48550/arXiv.2312.02566}
